Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2017-06-08 latex/deforestmap.tex
Add latex

\documentclass[a4paper, 12pt, leqno]{article} %leqno: numéro d'équation à gauche
\pagenumbering{arabic} % choose how to number the pages
\usepackage{a4wide}
\usepackage[utf8]{inputenc} % accents interprétés
\usepackage{graphicx}
%\usepackage{subfig}
%\usepackage[hmargin=2cm, vmargin = 2cm, noheadfoot]{geometry} %% Pour gérer le format des pages
%\usepackage{layout} %% Pour avoir la longueur des marges
\usepackage[round,sort]{natbib} %% Natbib is a popular style for formatting references.
%\usepackage{multibib}
%\newcites{secnm}{Bibliographie} 
%\usepackage{verbatim} % for multiline comments
\usepackage{amssymb} %symbole de maths
\usepackage{amsmath} %idem
%\usepackage{stmaryrd} %% Symbole flèche à l'envers
%\usepackage{amsfonts}
\usepackage[english]{babel} %% Les titres en anglais
\usepackage{array} %% Pour centrer verticalement le contenu d'un tableau, entre autres...
\setcounter{secnumdepth}{4} %% Profondeur des sections, subsections
\usepackage{setspace} %% Gère l'interligne: singlespacing, doublespacing
\usepackage{booktabs}
%\singlespacing
\usepackage{longtable}
\setlength{\LTleft}{-5cm plus 1 fill}
\setlength{\LTright}{-5cm plus 1 fill}
\usepackage[colorlinks=true,citecolor=blue]{hyperref} %% Gère les hyperliens
\usepackage{lineno} %% Numérotation des lignes
\usepackage{caption2}
%\linenumbers
\newcommand{\logit}{\text{logit}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\p}{\text{p}}
\newcommand{\R}{\textnormal{\sffamily\bfseries R}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

@@

  2. In this study, we propose a new approach combining historical
  (1953-2000) national forest-cover maps with recent (2000-2014)
  global annual tree cover loss data to look at six decades
  (1953-2014) of deforestation and forest fragmentation in
  Madagascar. We produced new forest-cover maps at 30m-resolution over
  the full territory of Madagascar for the year 1990, and annually
  from 2000 to 2014.

  3. We estimated that Madagascar has lost 44\% of its natural forest
  cover over the period 1953-2014 (including 37\% over the period
  1973-2014).  Natural forests cover 8.9 Mha in 2014 (15\% of the
  national territory) which are divided into 4.4 Mha (50\%) of moist
  forests, 2.6 Mha (29\%) of dry forests, 1.7 Mha of spiny forests
  (19\%) and 177,000 ha (2\%) of mangroves. Since 2005, the annual
  deforestation rate has progressively increased in Madagascar to
  reach 99,000 ha/yr during 2010-2014 (corresponding to a rate of
  1.08\%/yr). Around half of the forest (46\%) is now located at less
  than 100m from the forest edge.

  4. \emph{Policy implications}: Increase in the deforestation rate
  since 2005 can be related to rapid population growth (close to
  3\%/yr) and to poor law enforcement due to political instability in
  the country.  More effort, including conservation and development
  programs, is needed in Madagascar to protect the remaining natural
  tropical forests, both to enhance local people's livelihoods and to
  support global environmental efforts.

@@

\keywords{biodiversity, climate-change, deforestation, Landsat,
  Madagascar, tropical forest}

@@

Separated from the African continent and the Indian plate about 165
and 88 million years ago respectively \citep{Ali2008}, the flora and
fauna of Madagascar followed its own evolutionary path. Isolation
combined with a high number of micro-habitats \citep{Pearson2009} has
led to Madagascar's exceptional biodiversity both in term of number of
species and endemism in many taxonomic groups \citep{Crottini2012,
  Goodman2005}.  Most of the biodiversity in Madagascar is
concentrated in the tropical forests of the island which can be
divided into three types: the moist forest in the East, the dry forest
in the West and the spiny forest in the South
\citep{Vieilledent2016}. This unparalleled biodiversity is severely
threatened by anthropogenic deforestation \citep{Harper2007,
  Vieilledent2013} associated with activities such as slash-and-burn
agriculture and pasture \citep{Scales2011}. Tropical forests in
Madagascar also store a large amount of carbon \citep{Vieilledent2016}
and high rates of deforestation in Madagascar are responsible for
large CO2 emissions in the atmosphere
\citep{Achard2014}. Deforestation threatens species survival by
directly reducing their available habitat \citep{Brooks2002,
  Tidd2001}. Forest fragmentation can also lead to species extinction
by isolating populations from each other and creating forest patches
too small to maintain viable populations
\citep{Saunders1991}. Fragmentation also increases forest edge where
ecological conditions (such as air temperature, light intensity and
air moisture) can be dramatically modified, with consequences on the
abundance and distribution of species \citep{Murcia1995}. Forest
fragmentation can also have substantial effects on forest carbon
storage capacity, as carbon stocks are much lower at the forest edge
than under a closed canopy \citep{Brinck2017}.  Moreover, forest
carbon stocks vary spatially due to climate or soil factors
\citep{Saatchi2011, Vieilledent2016}. As a consequence, accurate and
spatially explicit maps of forest-cover and forest-cover change are
necessary to monitor biodiversity loss and carbon emissions from
deforestation and forest fragmentation, assess the efficiency of
present conservation strategies \citep{Eklund2016}, and implement new
strategies for the future \citep{Vieilledent2013,
  Vieilledent2016}. Simple time-series of forest-cover estimates, such
as those provided by the FAO Forest Resource Assessment report
\citep{Keenan2015} are not sufficient.

Unfortunately, accurate and exhaustive forest-cover maps are not
available for Madagascar for the last fifteen years (2000-2015).
\citet{Harper2007} produced maps of forest cover and forest cover
changes over Madagascar for the years c. 1953, c. 1973, 1990 and 2000.
The c. 1953 forest map was derived from the visual interpretation of
aerial photography at coarse scale (1/1,000,000). Forest maps for the
years c. 1973, 1990 and 2000, were obtained from supervised
classification of Landsat satellite images at 60m resolution (for the
year 1973) or 30m resolution (for years 1990 and 2000) and can be used
to derive accurate estimates (89.5\% accuracy reported for the forest
/ non-forest map of year 2000). Maps provided by \citet{Harper2007}
were not exhaustive (due to the presence of clouds in the satellite
imagery), e.g.~11 244 km2 are mapped as unknown cover type for the
year 2000.  Using a similar supervised classification approach as in
\citet{Harper2007}, more recent maps have been produced for the
periods 2000-2005-2010 by national institutions, with the technical
support of international environmental NGOs \citep{MEFT2009,
  ONE2013}. Another set of recent forest-cover maps using an advanced
statistical tool for classification, the Random Forest classifier
\citep{Grinand2013, Rakotomalala2015}, was produced for the periods
2005-2010-2013 \citep{ONE2015}. However, these maps are either too old
to give recent estimates of deforestation \citep{MEFT2009, ONE2013},
include large areas of missing information (due to images with more
than 20\% of clouds) for the 2010 map provided by \citep{ONE2013}, or
show large mis-classification in specific areas, especially in the dry
and spiny forest domain for which the spectral answer has a strong
seasonal behavior due to the deciduousness of such forests (overall
accuracy is lower than 0.8 for the dry and spiny forests for the maps
produced by \citet{ONE2015}). Moreover, the production of such forest
maps from a supervised classification approach requires significant
resources, especially regarding the image selection step (required to
minimize cloud cover) and the training step (visual interpretation of
a large number of polygons needed to train the classification
algorithm) \citep{Rakotomalala2015}. Most of this work of image
selection and visual interpretation would need to be repeated to
produce new forest maps in the future using a similar approach.

Global forest or tree cover products have also been published recently
and can be tested at the national scale for
Madagascar. \citep{Kim2014} produced a global forest-cover change map
from 1990 to 2000 (derived from Landsat imagery). This product was
updated to cover the period 1975-2005
(\url{http://glcf.umd.edu/data/landsatFCC/}) but forest-cover maps
after 2005 were not produced. Moreover, the approach used in
\citet{Kim2014} did not accurately map the forests in the dry and
spiny ecosystems of Madagascar (see Fig. 8 in \citet{Kim2014}).
\citet{Hansen2013} mapped tree cover percentage, annual forest loss
and forest gain from 2000 to 2012 at global scale at 30 m
resolution. This product has since been updated and is now available
up to the year 2014 \citep{Hansen2013}. To map forest cover from the
\citet{Hansen2013} product, a tree cover threshold must be selected
(that defines forest cover). Selecting such a threshold is not
straightforward as the accuracy of the global tree cover map strongly
varies between forest types, and is substantially lower for dry
forests than for moist forests \citep{Bastin2017}. Moreover, the
\citet{Hansen2013} product does not provide information on
land-use. In particular the global tree cover map does not separate
tree plantations such as oil palm or eucalyptus plantations from
natural forests \citep{Tropek2014}. Thus, the global tree cover map
from \citet{Hansen2013} cannot be used alone to produce a map of
forest cover \citep{Tyukavina2017}. In complement to the tree cover
percentage provided in \citet{Hansen2013}, a layer of annual tree
cover loss is also provided (i.e.~complete loss of tree cover from a
value higher than 10\% to zero) for the period 2001-2014.

In this study, we present a simple approach which combines the maps
from \citet{Harper2007} and products from \citet{Hansen2013} to derive
annual wall-to-wall forest-cover change maps over the period 2000-2014
for Madagascar. We use the forest-cover map provided by
\citet{Harper2007} for the year 2000 (defining the land-use) with the
tree cover loss product provided by \citet{Hansen2013} that we apply
only inside forest areas identified by \citet{Harper2007}. Similar to
the approach of \citet{Harper2007}, we also assess trends in
deforestation rates and forest fragmentation from c. 1953 to 2014. The
approach described in this study can help assess the effectiveness of
current conservation strategies, and assist the implementation of
future strategies. Our approach could be easily extended to other
tropical countries that have at least one forest-cover map between
2000 and 2014. This approach can easily be repeated in the future when
the \citet{Hansen2013} products are updated.

@@

We produced annual forest/non-forest maps at 30m resolution for the
full territory of Madagascar for the period 2000-2014 by combining the
forest map of year 2000 from \citet{Harper2007}, and the tree cover
percentage and annual forest cover loss maps over the period 2000-2014
from \citet{Hansen2013}. The 2000 Harper's forest map includes 208,000
ha of unclassified areas due to the presence of clouds in input
satellite imagery, mostly (88\%) within the moist forest domain which
covers 4.17 million ha in total. To provide a label (forest or
non-forest) to these unclassified pixels, we use the 2000 tree cover
percentage map of \citet{Hansen2013} by selecting a threshold of 75\%
tree cover to define forest cover as recommended by other studies for
the moist domain \citep{Achard2014}. We thus obtained a forest-cover
map for the year 2000 covering the full territory of Madagascar. We
then combined this forest-cover map of the year 2000 with the annual
forest cover loss maps from 2001 to 2014 provided by
\citet{Hansen2013} to create annual forest-cover maps from 2001 to
2014 at 30m resolution. We also completed the Harper's forest map of
year 1990 by filling unclassified areas (due to the presence of clouds
on satellite images) using our forest-cover map of year 2000. To do
so, we assumed that if forest was present in 2000, the pixel was also
forested in 1990. The remaining unclassified pixels were limited to a
relatively small total area of c. 8,000 ha. We labeled these residual
pixels as non-forest as for the year 2000.  Similarly we completed the
Harper forest map of year 1973 by filling unclassified areas using our
forest-cover map of the year 1990 assuming that if forest was present
in 1990, it was also present in 1973.  Contrary to the year 1990, the
remaining unclassified pixels for year 1973 corresponded to a
significant total area of 3.32 million ha. We also reprojected the
forest-cover map of year 1953 to a common projection in order to
compare to the other eras. This map was produced by scanning a paper
map derived from aerial photos, and thus could not be perfectly
aligned with the other maps produced through digital processing of
satellite imagery \citep{Harper2007}. Finally for all forest-cover
maps from 1973, the isolated single non-forest pixels (i.e.~fully
surrounded by forest pixels) were removed, assuming that single
non-forest pixels inside a forest patch were not corresponding to
deforestation (they might correspond to selective logging activities).
This allowed us to avoid counting very small scale events
(\textless{}0.1 ha such as selective logging) as forest fragmentation.
All the resulting maps are freely available at
\url{https://bioscenemada.cirad.fr/forestmaps}.

@@

\begin{equation}
  \theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}
\end{equation}

@@

We also conducted an analysis of changes in forest fragmentation for
the years 1953, 1973, 1990, 2000, 2010 and 2014. We applied the method
developed by \citet{Riitters2000} which uses a moving window to
characterize the fragmentation around each forested
pixel. Computations were done using the function \texttt{r.forestfrag}
of the GRASS GIS software \citep{Neteler2008}. Six categories of
fragmentation were identified from the amount of forest and its
occurrence as adjacent forest pixels: ``interior'', ``perforated'',
``edge'', ``transitional'', ``patch'', and ``undetermined''. We used a
moving window of 7x7 pixels (4.4 ha). Using this window size, forest
edge had a width of about 90m \citep{Riitters2000}. The ``interior''
category can be interpreted as the most intact forest
\citep{Potapov2017}. The ``patch'' and ``transitional'' categories
correspond to isolated small forest patches.  We reported the area of
forest in each fragmentation category for the six years and analyzed
the dynamics of fragmentation over the six decades. We also computed
the distance to forest edge for all forest pixels for the years 1953,
1973, 1990, 2000, 2005, 2010 and 2014. For that, we used the function
\texttt{gdal\_proximity.py} of the GDAL software
(\url{http://www.gdal.org/}). We computed the mean and 95\% quantiles
of the distance to forest edge and looked at the evolution of these
values with time.

@@

Natural forests in Madagascar covered 16.0 Mha in 1953, about 27\% of
the national territory of 587,041 km2. In 2014, the forest cover
dropped to 8.9 Mha, corresponding to about 15\% of the national
territory (Fig.~\ref{fig:fcc} and
Tab.~\ref{tab:forest_cover}). Madagascar has lost 44\% and 37\% of its
natural forests between 1953 and 2013, and between c. 1973 and 2014
respectively (Fig.~\ref{fig:fcc} and Tab.~\ref{tab:forest_cover}). In
2014 the remaining 8.9 Mha of natural forest were distributed as: 4.4
Mha of moist forest (50\% of total forest cover), 2.6 Mha of dry
forest (29\%), 1.7 Mha of spiny forest (19\%) and 0.18 Mha (2\%) of
mangrove forest (Fig.~\ref{fig:ecoregion} and
Tab.~\ref{tab:comp_forest}). Regarding the deforestation trend, we
observed a progressive decrease of the deforestation rate after1990
from 205,000 ha/yr (1.63\%/yr) over the period 1973-1990 to 44,300
ha/yr (0.43\%/yr) over the period 2000-2005
(Tab.~\ref{tab:forest_cover}). Then from 2005, the deforestation rate
has progressively increased and has more than doubled over the period
2010-2014 (98,700 ha/yr, 1.08\%/yr) compared to 2000-2005
(Tab.~\ref{tab:forest_cover}). The deforestation trend characterized
by a progressive decrease of the deforestation rate over the period
1990-2005 and a progressive increase of the deforestation after 2005
is valid for all four ecoregions (Tab.~\ref{tab:comp_defor}), with the
exception of the spiny forest domain for which the deforestation rate
during the period 2010-2013 was lower than during 2005-2010
(Tab.~\ref{tab:comp_defor}).

@@

Forest-cover maps provided by previous studies over Madagascar were
not exhaustive (unclassified areas) due to the presence of clouds in
satellite imagery used to produce such maps. In \citet{Harper2007},
the maps of years 1990 and 2000 include 0.5 and 1.12 Mha of unknown
cover type respectively. Proportions of unclassified areas are not
reported in the two other existing studies by \citet{MEFT2009} and
\citet{ONE2015}.  With our approach, we produced wall to wall
forest-cover change maps from 1990 to 2014 for the full territory of
Madagascar (Tab.~\ref{tab:forest_cover}). This allowed us to produce
more robust estimates of forest-cover and deforestation rates over
this period. Our forest-cover estimates over the period 1953-2013
(considering forest cover estimates at national level and by
ecoregions for the 6 available dates) were well correlated (0.99) to
estimates from the three previous studies (Tab.~\ref{tab:comp_forest})
with a RMSE of 300,000 ha (6\% of the mean forest cover of 4.8 Mha
when considering all dates and forest types together). These small
differences can be partly attributed to differences in ecoregion
boundaries. Despite significant differences in deforestation estimates
(Tab.~\ref{tab:comp_defor}), a similar deforestation trend was
observed across studies with a decrease of deforestation rates over
the period 1990-2005, followed by a progressive increase of the
deforestation after 2005.

@@

In parallel to the dynamics of deforestation, forest fragmentation has
progressively increased since 1953 in Madagascar. We observed a
continuous decrease of the mean distance to forest edge from 1953 to
2014 in Madagascar. The mean distance to forest edge has decreased to
c.  300 m in 2014 while it was previously c. 1.5 km in 1973
(Fig.~\ref{fig:dist_edge}).  Moreover, a large proportion (73\%) of
the forest was located at a distance greater than 100 m in 1973, while
almost half of the forest (46\%) was at a distance lower than 100 m
from forest edge in 2014 (Fig.  3). The percentage of forest that can
be considered intact in Madagascar has continuously decreased since
1953. The percentage of forest belonging to the ``interior'' category
(most intact forests) has fallen from 68\% in 1973 to 50\% in 2014. In
2014, more than 16\% of the forest belonged to the ``patch'' and
``transitional'' categories (isolated small forest patches) compared
to 9.5\% in 1973 (Tab.~\ref{tab:frag}).

@@

\bibliographystyle{gcb}
\bibliography{bib/biblio}

>>>>

\documentclass[a4paper, 12pt, leqno]{article} %leqno: numéro d'équation à gauche
\pagenumbering{arabic} % choose how to number the pages
\usepackage{a4wide}
\usepackage[utf8]{inputenc} % accents interprétés
\usepackage{graphicx}
%\usepackage{subfig}
%\usepackage[hmargin=2cm, vmargin = 2cm, noheadfoot]{geometry} %% Pour gérer le format des pages
%\usepackage{layout} %% Pour avoir la longueur des marges
\usepackage[round,sort]{natbib} %% Natbib is a popular style for formatting references.
%\usepackage{multibib}
%\newcites{secnm}{Bibliographie} 
%\usepackage{verbatim} % for multiline comments
\usepackage{amssymb} %symbole de maths
\usepackage{amsmath} %idem
%\usepackage{stmaryrd} %% Symbole flèche à l'envers
%\usepackage{amsfonts}
\usepackage[english]{babel} %% Les titres en anglais
\usepackage{array} %% Pour centrer verticalement le contenu d'un tableau, entre autres...
\setcounter{secnumdepth}{4} %% Profondeur des sections, subsections
\usepackage{setspace} %% Gère l'interligne: singlespacing, doublespacing
\usepackage{booktabs}
%\singlespacing
\usepackage{longtable}
\setlength{\LTleft}{-5cm plus 1 fill}
\setlength{\LTright}{-5cm plus 1 fill}
\usepackage[colorlinks=true,citecolor=blue]{hyperref} %% Gère les hyperliens
\usepackage{lineno} %% Numérotation des lignes
\usepackage{caption}
%\linenumbers
\newcommand{\logit}{\text{logit}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\p}{\text{p}}
\newcommand{\R}{\textnormal{\sffamily\bfseries R}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

@@

  2. In this study, we propose a new approach combining historical
  (1953-2000) national forest-cover maps with recent (2001-2014)
  global annual tree cover loss data to look at six decades
  (1953-2014) of deforestation and forest fragmentation in
  Madagascar. We produced new forest-cover maps at 30~m resolution over
  the full territory of Madagascar for the year 1990, and annually
  from 2000 to 2014.

  3. We estimated that Madagascar has lost 44\% of its natural forest
  cover over the period 1953-2014 (including 37\% over the period
  1973-2014). Natural forests cover 8.9 Mha in 2014 (15\% of the
  national territory) which are divided into 4.4 Mha (50\%) of moist
  forests, 2.6 Mha (29\%) of dry forests, 1.7 Mha of spiny forests
  (19\%) and 177,000 ha (2\%) of mangroves. Since 2005, the annual
  deforestation rate has progressively increased in Madagascar to
  reach 99,000 ha/yr during 2010-2014 (corresponding to a rate of
  1.08\%/yr). This increase is probably due to rapid population growth
  (close to 3\%/yr) and to poor law enforcement in the country. Around
  half of the forest (46\%) is now located at less than 100m from the
  forest edge.

  4. \emph{Policy implications}: Accurate forest-cover change maps can
  be used to assess the effectiveness of past and current conservation
  programs and implement new strategies for the future. In particular,
  forest maps and estimates can be used in the framework of the REDD+
  (``Reducing Emissions from Deforestation and Forest Degradation'')
  initiative and for optimizing the current protected area network.

@@

\keywords{biodiversity, climate-change, deforestation, Madagascar, tropical forest}

@@

Separated from the African continent and the Indian plate about 165
and 88 million years ago respectively \citep{Ali2008}, the flora and
fauna of Madagascar followed its own evolutionary path. Isolation
combined with a high number of micro-habitats \citep{Pearson2009} has
led to Madagascar's exceptional biodiversity both in term of number of
species and endemism in many taxonomic groups \citep{Crottini2012,
  Goodman2005}. Most of the biodiversity in Madagascar is
concentrated in the tropical forests of the island which can be
divided into four types: the moist forest in the East, the dry forest
in the West, the spiny forest in the South and the mangroves on the
West coast \citep{Vieilledent2016}. This unparalleled biodiversity is
severely threatened by deforestation \citep{Harper2007,
  Vieilledent2013} associated with human activities such as slash-and-burn
agriculture and pasture \citep{Scales2011}. Tropical forests in
Madagascar also store a large amount of carbon \citep{Vieilledent2016}
and high rates of deforestation in Madagascar are responsible for
large CO$_2$ emissions in the atmosphere
\citep{Achard2014}. Deforestation threatens species survival by
directly reducing their available habitat \citep{Brooks2002,
  Tidd2001}. Forest fragmentation can also lead to species extinction
by isolating populations from each other and creating forest patches
too small to maintain viable populations
\citep{Saunders1991}. Fragmentation also increases forest edge where
ecological conditions (such as air temperature, light intensity and
air moisture) can be dramatically modified, with consequences on the
abundance and distribution of species \citep{Murcia1995}. Forest
fragmentation can also have substantial effects on forest carbon
storage capacity, as carbon stocks are much lower at the forest edge
than under a closed canopy \citep{Brinck2017}. Moreover, forest
carbon stocks vary spatially due to climate or soil factors
\citep{Saatchi2011, Vieilledent2016}. As a consequence, accurate and
spatially explicit maps of forest-cover and forest-cover change are
necessary to monitor biodiversity loss and carbon emissions from
deforestation and forest fragmentation, assess the efficiency of
present conservation strategies \citep{Eklund2016}, and implement new
strategies for the future \citep{Vieilledent2013,
  Vieilledent2016}. Simple time-series of forest-cover estimates, such
as those provided by the FAO Forest Resource Assessment report
\citep{Keenan2015} are not sufficient.

Unfortunately, accurate and exhaustive forest-cover maps are not
available for Madagascar for the last fifteen years (2000-2015).
\citet{Harper2007} produced maps of forest cover and forest cover
changes over Madagascar for the years \emph{c.}~1953, \emph{c.}~1973,
1990 and 2000. The \emph{c.}~1953 forest map was derived from the
visual interpretation of aerial photography at coarse scale
(1/1,000,000). Forest maps for the years \emph{c.}~1973, 1990, and
2000 were obtained from supervised classification of Landsat satellite
images at 60~m resolution (for the year 1973) or 30~m resolution (for
years 1990 and 2000) and can be used to derive more accurate estimates
of forest cover (89.5\% accuracy reported for the forest/non-forest
map of year 2000). Nonetheless, maps provided by \citet{Harper2007}
are not exhaustive (due to the presence of clouds in the satellite
imagery), e.g.~11 244~km2 are mapped as unknown cover type for the
year 2000. Using a similar supervised classification approach as in
\citet{Harper2007}, more recent maps have been produced for the
periods 2000-2005-2010 by national institutions, with the technical
support of international environmental NGOs \citep{MEFT2009,
  ONE2013}. Another set of recent forest-cover maps using an advanced
statistical tool for classification, the Random Forest classifier
\citep{Grinand2013, Rakotomalala2015}, was produced for the periods
2005-2010-2013 \citep{ONE2015}. However, these maps are either too old
to give recent estimates of deforestation \citep{MEFT2009, ONE2013},
include large areas of missing information due to images with high
percentage of cloud cover \citep{ONE2013}, or show large
mis-classification in specific areas, especially in the dry and spiny
forest domain for which the spectral answer has a strong seasonal
behavior due to the deciduousness of such forests (overall accuracy is
lower than 0.8 for the dry and spiny forests for the maps produced by
\citet{ONE2015}). Moreover, the production of such forest maps from a
supervised classification approach requires significant resources,
especially regarding the image selection step (required to minimize
cloud cover) and the training step (visual interpretation of a large
number of polygons needed to train the classification algorithm)
\citep{Rakotomalala2015}. Most of this work of image selection and
visual interpretation would need to be repeated to produce new forest
maps in the future using a similar approach.

Global forest or tree cover products have also been published recently
and can be tested at the national scale for
Madagascar. \citet{Kim2014} produced a global forest-cover change map
from 1990 to 2000 (derived from Landsat imagery). This product was
updated to cover the period 1975-2005
(\url{http://glcf.umd.edu/data/landsatFCC/}) but forest-cover maps
after 2005 were not produced. Moreover, the approach used in
\citet{Kim2014} did not accurately map the forests in the dry and
spiny ecosystems of Madagascar (see Fig. 8 in \citet{Kim2014}).
\citet{Hansen2013} mapped tree cover percentage, annual forest loss
and forest gain from 2000 to 2012 at global scale at 30 m
resolution. This product has since been updated and is now available
up to the year 2014 \citep{Hansen2013}. To map forest cover from the
\citet{Hansen2013} product, a tree cover threshold must be selected
(that defines forest cover). Selecting such a threshold is not
straightforward as the accuracy of the global tree cover map strongly
varies between forest types, and is substantially lower for dry
forests than for moist forests \citep{Bastin2017}. Moreover, the
\citet{Hansen2013} product does not provide information on
land-use. In particular the global tree cover map does not separate
tree plantations such as oil palm or eucalyptus plantations from
natural forests \citep{Tropek2014}. Thus, the global tree cover map
from \citet{Hansen2013} cannot be used alone to produce a map of
forest cover \citep{Tyukavina2017}. In complement to the tree cover
percentage provided in \citet{Hansen2013}, a layer of annual tree
cover loss is also provided (i.e.~complete loss of tree cover from a
value higher than 10\% to zero) for the period 2001-2014.

In this study, we present a simple approach which combines the maps
from \citet{Harper2007} and products from \citet{Hansen2013} to derive
annual wall-to-wall forest-cover change maps over the period 2000-2014
for Madagascar. We use the forest-cover map provided by
\citet{Harper2007} for the year 2000 (defining the land-use) with the
tree cover loss product provided by \citet{Hansen2013} that we apply
only inside forest areas identified by \citet{Harper2007}. Similar to
the approach of \citet{Harper2007}, we also assess trends in
deforestation rates and forest fragmentation from \emph{c.} 1953 to 2014. The
approach described in this study can help assess the effectiveness of
current conservation strategies, and assist the implementation of
future strategies. Our approach could be easily extended to other
tropical countries that have at least one forest-cover map between
2000 and 2014. This approach can easily be repeated in the future when
the \citet{Hansen2013} products are updated.

@@

We produced annual forest/non-forest maps at 30~m resolution for the
full territory of Madagascar for the period 2000-2014 by combining the
forest map of year 2000 from \citet{Harper2007}, and the tree cover
percentage and annual forest cover loss maps over the period 2000-2014
from \citet{Hansen2013}. The 2000 Harper's forest map includes 208,000
ha of unclassified areas due to the presence of clouds on satellite
images, mostly (88\%) within the moist forest domain which covered
4.17 Mha in total in 2000. To provide a label (forest or non-forest)
to these unclassified pixels, we used the 2000 tree cover percentage
map of \citet{Hansen2013} by selecting a threshold of 75\% tree cover
to define forest cover as recommended by other studies for the moist
domain \citep{Achard2014}. We thus obtained a forest-cover map for the
year 2000 covering the full territory of Madagascar. We then combined
this forest-cover map of the year 2000 with the annual tree cover loss
maps from 2001 to 2014 provided by \citet{Hansen2013} to create annual
forest-cover maps from 2001 to 2014 at 30~m resolution. We also
completed the Harper's forest map of year 1990 by filling unclassified
areas (due to the presence of clouds on satellite images) using our
forest-cover map of year 2000. To do so, we assumed that if forest was
present in 2000, the pixel was also forested in 1990. The remaining
unclassified pixels were limited to a relatively small total area of
\emph{c.} 8,000 ha. We labeled these residual pixels as non-forest as
for the year 2000. Similarly we completed the Harper's forest map of
year 1973 by filling unclassified areas using our forest-cover map of
the year 1990 assuming that if forest was present in 1990, it was also
present in 1973. Contrary to the year 1990, the remaining unclassified
pixels for year 1973 corresponded to a significant total area of 3.32
million ha. We also reprojected the forest-cover map of year 1953 to a
common projection in order to compare the forest-cover area in 1953
with forest-cover areas at the following dates. This map was produced
by scanning a paper map derived from aerial photos, and thus could not
be perfectly aligned with the other maps produced through digital
processing of satellite imagery \citep{Harper2007}. Finally for all
forest-cover maps from 1973, the isolated single non-forest pixels
(i.e.~fully surrounded by forest pixels) were removed, assuming that
single non-forest pixels inside a forest patch were not corresponding
to deforestation (they might correspond to selective logging
activities). This allowed us to avoid counting very small scale events
(\textless{}0.1 ha such as selective logging) as forest
fragmentation. All the resulting maps are freely available at
\url{https://bioscenemada.cirad.fr/forestmaps}.

@@

\begin{equation*}
  \theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}
\end{equation*}

@@

We also conducted an analysis of changes in forest fragmentation for
the years 1953, 1973, 1990, 2000, 2005, 2010 and 2014. We applied the
method developed by \citet{Riitters2000} which uses a moving window to
characterize the fragmentation around each forested
pixel. Computations were done using the function \texttt{r.forestfrag}
of the GRASS GIS software \citep{Neteler2008}. Six categories of
fragmentation were identified from the amount of forest and its
occurrence as adjacent forest pixels: ``interior'', ``perforated'',
``edge'', ``transitional'', ``patch'', and ``undetermined''. We used a
moving window of 7x7 pixels (4.4 ha). Using this window size, forest
edge had a width of about 90m \citep{Riitters2000}. The ``interior''
category can be interpreted as the most intact forest
\citep{Potapov2017}. The ``patch'' and ``transitional'' categories
correspond to isolated small forest patches. We reported the area of
forest in each fragmentation category for the six years and analyzed
the dynamics of fragmentation over the six decades. We also computed
the distance to forest edge for all forest pixels for the years 1953,
1973, 1990, 2000, 2005, 2010 and 2014. For that, we used the function
\texttt{gdal\_proximity.py} of the GDAL software
(\url{http://www.gdal.org/}). We computed the mean and 90\% quantiles
(5\% and 95\%) of the distance to forest edge and looked at the
evolution of these values with time.

@@

Natural forests in Madagascar covered 16.0 Mha in 1953, about 27\% of
the national territory of 587,041 km2. In 2014, the forest cover
dropped to 8.9 Mha, corresponding to about 15\% of the national
territory (Fig.~\ref{fig:fcc} and
Tab.~\ref{tab:forest_cover}). Madagascar has lost 44\% and 37\% of its
natural forests between 1953 and 2014, and between 1973 and 2014
respectively (Fig.~\ref{fig:fcc} and Tab.~\ref{tab:forest_cover}). In
2014 the remaining 8.9 Mha of natural forest were distributed as: 4.4
Mha of moist forest (50\% of total forest cover), 2.6 Mha of dry
forest (29\%), 1.7 Mha of spiny forest (19\%) and 0.18 Mha (2\%) of
mangrove forest (Fig.~\ref{fig:ecoregion} and
Tab.~\ref{tab:comp_forest}). Regarding the deforestation trend, we
observed a progressive decrease of the deforestation rate after 1990
from 205,000 ha/yr (1.63\%/yr) over the period 1973-1990 to 44,300
ha/yr (0.43\%/yr) over the period 2000-2005
(Tab.~\ref{tab:forest_cover}). Then from 2005, the deforestation rate
has progressively increased and has more than doubled over the period
2010-2014 (98,700 ha/yr, 1.08\%/yr) compared to 2000-2005
(Tab.~\ref{tab:forest_cover}). The deforestation trend characterized
by a progressive decrease of the deforestation rate over the period
1990-2005 and a progressive increase of the deforestation after 2005
is valid for all four ecoregions (Tab.~\ref{tab:comp_defor}), with the
exception of the spiny forest domain for which the deforestation rate
during the period 2010-2013 was lower than during 2005-2010
(Tab.~\ref{tab:comp_defor}).

@@

Forest-cover maps provided by previous studies over Madagascar were
not exhaustive (unclassified areas) due to the presence of clouds on
satellite images used to produce such maps. In \citet{Harper2007}, the
maps of years 1990 and 2000 include 0.5 and 1.12 Mha of unknown cover
type respectively. Proportions of unclassified areas are not reported
in the two other existing studies by \citet{MEFT2009} and
\citet{ONE2015}. With our approach, we produced wall to wall
forest-cover change maps from 1990 to 2014 for the full territory of
Madagascar (Tab.~\ref{tab:forest_cover}). This allowed us to produce
more robust estimates of forest-cover and deforestation rates over
this period. Our forest-cover estimates over the period 1953-2013
(considering forest cover estimates at national level and by
ecoregions for all the available dates) were well correlated
(Pearson's correlation coefficient $=$ 0.99) to estimates from the
three previous studies (Tab.~\ref{tab:comp_forest}) with a RMSE of
300,000 ha (6\% of the mean forest cover of 4.8 Mha when considering
all dates and forest types together). These small differences can be
partly attributed to differences in ecoregion boundaries. Despite
significant differences in deforestation estimates
(Tab.~\ref{tab:comp_defor}), a similar deforestation trend was
observed across studies with a decrease of deforestation rates over
the period 1990-2005, followed by a progressive increase of the
deforestation after 2005.

@@

In parallel to the dynamics of deforestation, forest fragmentation has
progressively increased since 1953 in Madagascar. We observed a
continuous decrease of the mean distance to forest edge from 1953 to
2014 in Madagascar. The mean distance to forest edge has decreased to
\emph{c.} 300~m in 2014 while it was previously \emph{c.} 1.5~km in 1973
(Fig.~\ref{fig:dist_edge}). Moreover, a large proportion (73\%) of
the forest was located at a distance greater than 100 m in 1973, while
almost half of the forest (46\%) was at a distance lower than 100 m
from forest edge in 2014 (Fig.~\ref{fig:dist_edge}). The percentage of
forest that can be considered intact in Madagascar has continuously
decreased since 1953. The percentage of forest belonging to the
``interior'' category (most intact forests) has fallen from 68\% in
1973 to 50\% in 2014. In 2014, more than 16\% of the forest belonged
to the ``patch'' and ``transitional'' categories (isolated small
forest patches) compared to 9.5\% in 1973 (Tab.~\ref{tab:frag}).

@@

%%\bibliographystyle{bib/jae}
%%\bibliography{bib/biblio}

\section{References}

\renewcommand{\bibsection}{} %% Remove section name for bibliography
\begin{thebibliography}{55}
\providecommand{\natexlab}[1]{#1}
